### 2017-10-21 src/2_Finance.tex
correcting chapter 2

\begin{description}
\item[Fill or Kill] is an order that must be executed immediately or being
cancelled, no partial fulfillments are allowed. 
\item[Day] the order is only valid during the day.
\item[Good til canceled] a order is active until the investor decides to cancel
it or the trade is executed.
\end{description}

@@

EMH is based on rational human behaviour and its validity has been criticised
by psychologists and behavioural economists who argue that the EMH is based on
counterfactual assumptions regarding human behaviour, that is, rationality.
Recent advances in evolutionary psychology and the cognitive neurosciences may
be able to reconcile the EMH with behavioural anomalies. On the other hand, if
the EMH is true, the market really walks randomly and therefore there shouldn't
be any difference between experienced and novice traders. Kim Man Lui proved
the contrary in a controlled experiment \cite{man2013}.

>>>>

\begin{description}
\item[Fill or Kill] is an order that must be executed immediately or being
cancelled, no partial fulfilments are allowed. 
\item[Day] the order is only valid during the day.
\item[Good till cancelled] a order is active until the investor decides to cancel
it or the trade is executed.
\end{description}

@@

EMH is based on rational human behaviour and its validity has been criticised
by psychologists and behavioural economists who argue that the EMH is based on
counterfactual assumptions regarding human behaviour, that is, rationality.
Recent advances in evolutionary psychology and the cognitive neuroscience may
be able to reconcile the EMH with behavioural anomalies. On the other hand, if
the EMH is true, the market really walks randomly and therefore there shouldn't
be any difference between experienced and novice traders. Kim Man Lui proved
the contrary in a controlled experiment \cite{man2013}.

### 2017-10-21 src/3_TimeSeries.tex
correcting chapter 2

\section{Characteristics of Financial Time Series}
There are several characteristics and concepts of financial time series and here we will discuss the more important ones.
\subsection{Stylized facts of asset returns}
\label{sec:stylizedfacts}

There are several known features exhibited by financial instruments called
stylized facts, which have been empirically studied and some of them have been
documented only recently and accepted as truth. Stylized facts are usually
formulated in terms of qualitative properties of asset returns and may not be
precise enough to distinguish among different parametric models.
\cite{cont2001}.

Some stylized facts which are common to a wide set of financial assets
\cite{sewell2011} are:

@@

 \begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}
\item[Distribution] The distribution of returns is approximately symmetric and
has high kurtosis (i.e fat tails and a peaked centre compared with the normal
distribution). However, distribution of returns whose were obtained from higher
frequencies looks more like a normal distribution.
The returns distribution tails are larger than
 what is hypothesised by common data generation process (generally normal
 distribution assumption). In the markets, fat tails are an undesirable feature
 because of the additional risk they imply.  In the figure \ref{fig:returndist}
 is shown the SPY returns distribution based on daily dates from the period 1st
 July 1998 to 4th April 2013. The distribution was compared against the normal
 distribution which clearly doesn't fit the data.  
 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}
\item[Heterogeneity] despite the fact that financial returns are non-stationary,
stationary periods can be observed. Economist speaks in terms of a structural
break \cite{stock1994}, in machine learning this is known as drift which means
that the statistical properties of the target variable change over time
\cite{widmer1996}, \cite{tsymbal2004}.
\item[Non-Linearity] financial returns may be non-linear in mean and/or
non-linear in variance.
\item[Calendar effects] are also called seasonal effects and they are cyclical
anomalies in returns where the cycle is based on the calendar. Some of known
calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
effect, the January effect and Holiday effect. The most important calendar
anomalies are the January effect and the weekend effect. 
\end{description}

\subsection{Stationary }
A strictly stationary times series $y_t$ is one for which the probabilistic behaviour
of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
to that of the time shifted set, more precisely: \[ P\{y_{t_1} \leq
c_1,\dots,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,\dots,y_{t_L+h} \leq c_L\}
\quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
$c_1,\dots,c_L$ are constants.  This definition is too strong and difficult to
assess it from a single data set. The weak version of this definition imposes
conditions only on the two first two moments.

@@

In finance, a shock represents an unexpected change in a variable or a in its
error term in a particular time period. For stationary time series, shocks to
the system will gradually die away. That is, a shock during time $t$ will have a
smaller effect in time $t+1$, a smaller effect on $t+2$ and so on. For
non-stationary data, the persistence of shocks will always be infinite.

@@

\subsection{Spurious regression} \label{sec:spurious}
The use of non-stationary data can lead to spurious regressions. Spurious
regression dates back to Yule in 1926 \cite{yule1926}. If two stationary
variables are generated as independent random series and are trending over
time, when one of those variables is regressed on the other, they could have a
high $R^2$ even if the two are totally unrelated. So, if standard regression
techniques are applied to non-stationary data, it could look good under
standard measures but valueless \cite{brooks2002}.
The
regression measure $R^2$ gives high values in presence of spurious
relationships. This measure would means that the model explains all the
variability of the response data around its mean which is not true.

@@

\subsubsection{Realized volatility} is also known as historic volatility and
it is the actual variance in the price of a stock over time.
Realized volatility is measured in terms of the standard deviation
using the historical stock prices. It is commonly calculated based on
intraday price returns:
\begin{equation}
\label{eq:retintra}
r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
\end{equation}
\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
intraday sample $n=2,\dots,N$. Realized volatility is defined as:
\begin{equation}
\label{eq:rv}
    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
\end{equation}
\noindent where $N$ is the number of intraday samples and $T$ is the
number of days. 
In order to include overnight returns, Hansen and
Lunde~\cite{hansen+lunde2005} introduced a scaling version of
realized volatility using the following definitions:
\begin{eqnarray}
r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
\end{eqnarray}
\noindent where equation~(\ref{eq: ret 1}) represents overnight 
returns and the volatility as 
equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
sample at day $t$. The scaled realized volatility $\rho(t)$ is 
defined as:
\begin{eqnarray}
\label{eq:srv}
\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
\end{eqnarray}
Realized volatility has also been defined as the absolute value return or as
the mean of the sum of intraday squared returns at short intervals of time. The
majority of research carried out in the literature obtain the daily volatility
as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
However, it has been proven that this measurement noise is too high for
observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen
and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could
result in an inferior model being chosen as the best one. The realized
volatility, as calculated by the cumulative sum of squared intraday returns and
shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an
inferior model.   
\subsubsection{Implied volatility} not only can be extracted from returns
but it can also be derived from option or future pricing models.  The
volatility obtained corresponds to the market's prediction of future
volatility. In finance, an option is a derivative, that is, a contract which
gives the owner the right, but not the obligation to buy or sell an underlying
asset at a given price called strike price. An option can be executed at any
time before an expiration date previously defined no matter what price the
underlying asset has. For example, the Black-Scholes model~\cite{black1973}
determines the fair option value based on stock price, strike price, time to
option expiration, the interest rate and volatility. These are known or can be
easily obtained from the market, excepting by volatility which must be
estimated. However, rather than assuming a volatility a priori and computing
option prices from it, the model can be used to estimate volatility at given
prices, time to expiration and strike price. This obtained volatility is called
the implied volatility of an option. Additionally, some models obtain implied
volatility from futures (other derivative from prices). For instance, the
Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used
to determine future volatilities~\cite{hamidetal2004}. Higher implied
volatility is indicative of greater price fluctuation in either direction.
Implied volatility is found by determining the value which makes theoretical
prices equal to market prices. In this way volatility is ``implied'' by the
current market price of the stock.

@@

For trading strategies, the interest is centred in forecasting realized
volatility over the life of an option and to take advantage when this
volatility differs from the implied volatility. This is called volatility
arbitrage. For example, a trader will buy an option and hedge the underlying
asset if the implied volatility is under the realized volatility.

@@

In the existing literature, there are four main classes of asset
return volatility models: the general autoregressive conditional
heteroskedasticity (GARCH) models, the stochastic volatility (SV)
models, the realized volatility models and the machine learning based
models. A comparison of the first three models can be found
in~\cite{wei2012}.

@@

The realized volatility constructed from high frequency intraday returns gave
rise to the realized volatility models mainly because the realized volatility
series is much more homoskedastic and seems to be a long memory
process~\cite{andersonetal2003}. For realized volatility, the autoregressive
fractionally integrated moving average (ARFIMA) process emerged as a standard
model~\cite{chenetal2010} and many variations have been studied, but all of
them produce similar forecasting results to the ARFIMA(1,d,1)
model~\cite{koopmanetal2005}.

>>>>

\section{Characteristics of Financial Time Series}
There are several characteristics and concepts of financial time series and here we will discuss the more important ones.
\subsection{Stylised facts of asset returns}
\label{sec:stylisedfacts}

There are several known features exhibited by financial instruments called
stylised facts, which have been empirically studied and some of them have been
documented only recently and accepted as truth. Stylised facts are usually
formulated in terms of qualitative properties of asset returns and may not be
precise enough to distinguish among different parametric models.
\cite{cont2001}.

Some stylised facts which are common to a wide set of financial assets
\cite{sewell2011} are:

@@

 \begin{figure}[h]
 \centering
 \includegraphics[width=0.7\textwidth]{plots/spy_returns_acf.pdf}
 \caption{SPY returns ACF}
 \label{fig:returnacf}
\end{figure}
\item[Distribution] The distribution of returns is approximately symmetric and
has high kurtosis (i.e fat tails and a peaked centre compared with the normal
distribution). 
The returns distribution tails are larger than
 what is hypothesised by common data generation process (generally normal
 distribution assumption). In the markets, fat tails are an undesirable feature
 because of the additional risk they imply.  However, distribution of returns whose were obtained from higher
frequencies looks more like a normal distribution.
In the figure \ref{fig:returndist}
 is shown the SPY returns distribution based on daily dates from the period 1st
 July 1998 to 4th April 2013. The distribution was compared against the normal
 distribution which clearly doesn't fit the data.  
 \begin{figure}[h]
 \centering
 \includegraphics[scale=0.5]{plots/spy_returns_dist.pdf}
 % spy_returns_dist.pdf: 504x504 pixel, 72dpi, 17.78x17.78 cm, bb=0 0 504 504
 \caption{SPY returns distribution}
 \label{fig:returndist}
\end{figure}
\item[Heterogeneity] despite the fact that financial returns are non-stationary,
stationary periods can be observed. Economist speaks in terms of a structural
break \cite{stock1994}, in machine learning this is known as drift which means
that the statistical properties of the target variable change over time
\cite{widmer1996}, \cite{tsymbal2004}.
\item[Non-Linearity] financial returns may be non-linear in mean and/or
non-linear in variance.
\item[Calendar effects] are also called seasonal effects and they are cyclical
anomalies in returns where the cycle is based on the calendar. Some of known
calendar effects are: intraday effect, weekend effect, Monday effect, intramonth
effect, the January effect and Holiday effect. The most important calendar
anomalies are the January effect and the weekend effect. 
\end{description}

\subsection{Stationary }
A strictly stationary times series $y_t$ is one for which the probabilistic behaviour
of every collection of values $\{y_{t_1},y_{t_2},\dots,y_{t_L}\}$ is identical
to that of the time shifted set, more precisely: \[ P\{y_{t_1} \leq
c_1,\dots,y_{t_L} \leq c_L\} = P\{y_{t_1+h} \leq c_1,\dots,y_{t_L+h} \leq c_L\}
\quad \forall L \in \mathbb{N}, \forall h \in \mathbb{Z}\] \noindent where
$c_1,\dots,c_L$ are constants.  This definition is too strong and difficult to
assess from a single data set. The weak version of this definition imposes
conditions only on the first two moments.

@@

One of the consequences of stationary processes is how they recover from a shock. A shock represents an unexpected change in a variable or a in its
error term in a particular period of time. For stationary time series, shocks to
the system will gradually die away. That is, a shock during time $t$ will have a
smaller effect in time $t+1$, a smaller effect on $t+2$ and so on. For
non-stationary data, the persistence of shocks will always have permanent effects because is not a mean-reverting process.

@@

\subsection{Spurious regression} \label{sec:spurious}
The use of non-stationary data can lead to spurious regressions. Spurious
regression dates back to Yule in 1926 \cite{yule1926}. If two stationary
variables are generated as independent random series and are trending over
time, when one of those variables is regressed on the other, they could have a
high coefficient of determination ($R^2$) even if the two are totally unrelated. So, if standard regression
techniques are applied to non-stationary data, it could look good under
standard measures but valueless \cite{brooks2002}.
The
regression measure $R^2$ gives high values in presence of spurious
relationships. This measure would means that the model explains all the
variability of the response data around its mean which is not true.

@@

\subsubsection{Realised volatility} is also known as historic volatility and
it is the actual variance in the price of a stock over time.
Realised volatility is measured in terms of the standard deviation
using the historical stock prices. It is commonly calculated based on
intraday price returns:
\begin{equation}
\label{eq:retintra}
r_{t,n}=100(\ln(p_{t,n}) - \ln(p_{t,n-1}))
\end{equation}
\noindent where $p_{t,n}$ is the price observed at day $t=1,\dots,T$ and
intraday sample $n=2,\dots,N$. Realised volatility is defined as:
\begin{equation}
\label{eq:rv}
    \hat{\sigma}(t) = \sum_{n=1}^N r_{t,n}^2 \, , 
\end{equation}
\noindent where $N$ is the number of intraday samples and $T$ is the
number of days. 
In order to include overnight returns, Hansen and
Lunde~\cite{hansen+lunde2005} introduced a scaling version of
realised volatility using the following definitions:
\begin{eqnarray}
r_{t}&=&100(\ln(p_{t,N}) - \ln(p_{t-1,N})) \label{eq: ret 1} \\
\bar{\rho}(t) &=& \sum_{t=1}^T r_{t}^2  \label{eq: ret 2} \, .
\end{eqnarray}
\noindent where equation~(\ref{eq: ret 1}) represents overnight 
returns and the volatility as 
equation~(\ref{eq: ret 2}), where $p_{t,N}$ is the last intraday 
sample at day $t$. The scaled realised volatility $\rho(t)$ is 
defined as:
\begin{eqnarray}
\label{eq:srv}
\rho(t) = \gamma \hat{\rho}(t) \, , \qquad & \qquad \gamma = \displaystyle \frac{\bar{\rho}(t)}{\displaystyle\sum_{t=1}^T \hat{\rho}(t)}
\end{eqnarray}
Realised volatility has also been defined as the absolute value return or as
the mean of the sum of intraday squared returns at short intervals of time. The
majority of research carried out in the literature obtain the daily volatility
as the daily squared returns as is shown in equation~(\ref{eq: ret 2}).
However, it has been proven that this measurement noise is too high for
observing the true volatility process~\cite{andersen+bollerslev1998}. Hansen
and Lunde~\cite{hansen+lunde2006} stated that the use of a noisy proxy could
result in an inferior model being chosen as the best one. The realised
volatility, as calculated by the cumulative sum of squared intraday returns and
shown in equation (\ref{eq:rv}), is less noisy and doesn't lead to choosing an
inferior model.   
\subsubsection{Implied volatility} not only can be extracted from returns
but it can also be derived from option or future pricing models.  The
volatility obtained corresponds to the market's prediction of future
volatility. In finance, an option is a derivative, that is, a contract which
gives the owner the right, but not the obligation to buy or sell an underlying
asset at a given price called strike price. An option can be executed at any
time before an expiration date previously defined no matter what price the
underlying asset has. For example, the Black-Scholes model~\cite{black1973}
determines the fair option value based on stock price, strike price, time to
option expiration, the interest rate and volatility. These are known or can be
easily obtained from the market, excepting by volatility which must be
estimated. However, rather than assuming a volatility a priori and computing
option prices from it, the model can be used to estimate volatility at given
prices, time to expiration and strike price. This obtained volatility is called
the implied volatility of an option. Additionally, some models obtain implied
volatility from futures (other derivative from prices). For instance, the
Barone-Adesi and Whaley futures option model~\cite{baroneetal1987} is also used
to determine future volatilities~\cite{hamidetal2004}. Higher implied
volatility is indicative of greater price fluctuation in either direction.
Implied volatility is found by determining the value which makes theoretical
prices equal to market prices. In this way volatility is ``implied'' by the
current market price of the stock.

@@

For trading strategies, the interest is centred in forecasting realised
volatility over the life of an option and to take advantage when this
volatility differs from the implied volatility. This is called volatility
arbitrage. For example, a trader will buy an option and hedge the underlying
asset if the implied volatility is under the realised volatility.

@@

In the existing literature, there are four main classes of asset
return volatility models: the general autoregressive conditional
heteroskedasticity (GARCH) models, the stochastic volatility (SV)
models, the realised volatility models and the machine learning based
models. A comparison of the first three models can be found
in~\cite{wei2012}.

@@

The realised volatility constructed from high frequency intraday returns gave
rise to the realised volatility models mainly because the realised volatility
series is much more homoskedastic and seems to be a long memory
process~\cite{andersonetal2003}. For realised volatility, the autoregressive
fractionally integrated moving average (ARFIMA) process emerged as a standard
model~\cite{chenetal2010} and many variations have been studied, but all of
them produce similar forecasting results to the ARFIMA(1,d,1)
model~\cite{koopmanetal2005}.
